Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-08-07 chapters_altFormat/ch_data_analysis.tex
Put in Pawel and Vladimir edits

Three basic calibrations are performed on the raw data level within this task. The gain calibration calibrates the electronics will be described in Section\ref{sec:elecCalib}. The anode gain calibration which will be discussed in Section~\ref{sec:anodeCalib}, and the gating grid noise subtraction and time bucket cut which is described in Section~\ref{sec:ggnoisesub}.
%gain calibration 
%ggnoise subtraction
%GG time bucket cut

\subsection{Pulse Shape Algorithm (PSA) Task}
\label{sec:psa}
%Yoffset match the TPC-Vertex_y with BDC_Y
%YPedestalOffset what is it for?
%cobo gain matching
An algorithm loops over the raw time bucket spectrum to find all the unique pules in each pad. It is very likely that there are several pulses in a pad coming from multiple tracks passing under the same pad which are separated only in arrival time. Figure~\ref{fig:psaTask} shows the PSA task working on one particular pad in experimental data. The shaded histogram in every sub-figure represents the ADC raw spectrum of the pad. In Fig.~\ref{fig:psa0} we see the PSA has identified the first pulse and fitted it with the standard pulse; the result of the fit is shown by the red line. The fitted red line is then subtracted from the ADC spectrum, which is shown in the shaded histogram in Fig.~\ref{fig:psa1}. Here, the second pulse is found and fitted again with the same procedure; the red line will be subtracted from the raw ADC spectrum. The black line represents the sum of all the fitted pules in each step. This continues until all pules are found or until the PSA reaches the last bin in the spectrum. Figure~\ref{fig:psa4} shows the final result where 4 fitted pulses were found. The agreement between the total summed line and the raw ADC spectrum exemplifies how successful the PSA is in this particular set of data.

@@

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse. We define the start of the pulse to be the location in time when the rising edge of the pulse reaches of 10\% of its peak value. The height of the pulse is related to the charge induced on the pad, $Q$. Combined with the start time of the pulse $t_o$  it is saved into a new container representing a \emph{hit}, called the \emph{STHit} class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spacial coordinates, and charge information $Q$.

@@

  	\caption{Example of the pulse shape algorithm through each step of in processing a given pad. }
	\label{fig:psaTask}
\end{figure}

@@

 The function of the  helix track finding algorithm is to sort the hit cloud into unique sets of hits representing tracks. This is performed by a Riemann track finding algorithm where the Cartesian space is mapped onto a Riemann sphere. This is a particular transformation in which unique helices in the Cartesian space form great circles on the Riemann sphere. This is particularly useful since hits which correspond to a track form a helices the Cartesian space of the TPC coordinates. It is difficult to search for tracks in the hit cloud by fitting a general helix to different combinations of hits. It is much easier to search for collection of hits by fitting a plane to the great circles which are formed on the Riemann sphere. More details on the Riemann tracking can be found here \cite{spiritroot_paper}. From here the set of hits which form a unique track are stored into a new helix track container called the STHelixTrack class.

@@

\begin{figure}[!htb]
\centering
\includegraphics[scale=.5]{top_view_helix_ext.pdf}
\caption{Top down view of a fit to a track passing through several pads. Here we outline the pads in bold, where the charges $q_i$ represent the charged deposited on each pad. In this example, three clusters at the bottom are clustered in the x-direction and for the upper three clustered in the z-direction. The cluster representing each group of hits represent the average position of the track. The track fit represents the position of the avalanche, $\bar{x}$,  where the position from the center to each pad to the $\bar{x}$ position is defined as $\lambda_i$.}
\label{fig:topview}
\end{figure}

 The position along the clustering direction, $\overline{X}$, is calculated as the charge-weighted average, $\overline{X} = \sum_i q_ix_i$, where $q_i$ and $x_i$ are the charge and position of the i-th hit in the cluster. The other direction is set to the center of the pad. For example, if we are clustering hits along the x-axis, the z-position is set to the center of the pad in the z-direction and vice versa. Switching the clustering direction gives better position resolution than if we clustered only along one direction. You can imagine if we calculated the clusters only along the x-axis, tracks with $\theta \approx 0^{\circ}$ the x-position is not well defined. The cluster position and charge is stored into a new container representing the cluster called the STHitCluster class. All the clusters which belong to a particular helix track is also stored in the corresponding STHelixTrack class.

@@

\subsection{Momentum and track reconstruction (GENFIT)}
After all of the corrections are applied to the clusters, the cluster's position and charge values are then passed to a software package which reconstructs the momentum of the track called GENFIT. GENFIT utilizes a Kalman fitting algorithm which returns the estimated momentum and charge values for a given track \cite{genfit}. These values are then stored in into a new track container called the STRecoTrack class.

@@

\subsection{Gating grid noise subtraction}
\label{sec:ggnoisesub}
Opening the gating grid is essentially done by shorting the adjacent wires in the gating grid together, and allowing them to reach equilibrium as fast as possible. Due to coupling between the gating grid, the ground plane, and other effects, the impedance of the supply cables was not perfectly matched, causing an imbalance in the current where one side discharged faster than the other side. This caused an under-damped oscillating current in the grid, which created and  oscillating residual noise early in the time bucket spectrum. Figure~\ref{fig:ggNoiseSub} shows in the upper panel shows the ADC time bucket spectrum after averaging over 2000 events in a given pad. 
 %The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency.

@@

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule
& \multicolumn{3}{c}{$100 MeV Target$}\\
\cmidrule{2-4}
Particle &\phantom{abc} & Measured & Corrected & \% Difference Raw & \% Difference Corrected\\
\midrule
p   & 882.8 & 929.5 & 877.3   &  3.7  & -1.0  \\
d   & 817.1 & 831.15 & 797.94 &  1.7  & -2.3\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail for the calibration run taken with the Al target. }
\label{tb:cocktail100tar}
\end{table*}

@@

A cocktail of several light charged particles (p,d,t,${}^{3}$He,${}^{4}$He) of well defined magnetic rigidity was produced was produced and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer were set so that the momentum resolution of the beam was defined to better than $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied with no target in place. A thick Aluminum target was also inserted to provide a slightly lower energy calibration point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases the certain particles were not able to propagate or were strongly attenuated and resulted in measurements of poor statistical precision.

@@

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The energy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. We found that the momenta of light particles in the  cocktail calibration beam differed significantly (up to 5\% ) from the expected values. This is shown in Tables~cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. We could address this discrepancy by incorporating the smaller horizontal magnetic field components at the edge of the dipole and taking into account the electron drift velocities in the direction of $\vec{E}\times\vec{B}$ direction as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a more general way which also includes the correction for the presence of positive space charge in the field cage volume. The same correction technique was applied here, where the cocktail beam is the special case of zero space charge.

(Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. In general, all these light particles corrected momentum are of the order of 1\% smaller than value given by the rigidity selection in the BigRIPS separator. This discrepancy could reflect the accuracy of our knowledge of the magnetic field of the SAMURAI dipole. Without a definitive explanation for this discrepancy, we have not attempted further correction of the calculated momenta. We note that it is of the same order as the resolution of the measurement of these momenta.

@@

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires due to a leak in the gating grid that has since been corrected. Out of all the runs used in the analysis in this thesis -- see \ref{tb:runList} -- the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we artificially increase the gain of the pads which lie above these anode wires in the software. The multiplicative factor is estimated by plotting the energy loss in the high gain sections relative to the low gain sections.

@@

Figure~\ref{fig:highlowcal} shows the correlation plot of $\langle dE/dx\angle$ of the high versus the low gain sections. The effects of the high gain channel saturation can around a value of 400 ADC\si{\per\milli\metre} and plateaus, where as the low gain does not; this region is left out of the fit. A linear fit was performed for the calibration between the high gain channel $h_c = G\cdot l_c$, where $l_c$ is the value of the low gain channel and $G$ is the gain factor. In the case where the low anode voltage was \SI{1214}{\volt}, the $G=9.8$, where in the case of \SI{1085}{\volt}, the $G = $.

 A map defining the gain calibration of each pad is pre-loaded into the software. In the decoder task the raw ADC spectra was multiplied by this gain factor for each channel. The PSA thresholds of those channels also were multiplied by the same gain factor since the noise levels were also artificially magnified as well.

@@

Using a TPC for measurements of HICs in intermediate energy heavy ion collisions presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments fundamental particles are produced with charge $\pm e$. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum ionization in the energy loss. In these cases the dynamic range of most TPC electronics can cover a wide range of particles. In nuclear HICs, we are interested in measured particles with charges $Ze$ where Z=1-3, and even higher in some applications. Since the energy loss in the Bethe-Bloch equation, Eq.~\ref{eq:bb}, is proportional to $Z^2$, the range of energy losses reflects the possible range of z values. Even more, in low energy nuclear HIC of intermediate beam energies -- around \SI{300}{\MeVA} -- produce low velocity particles which exist in the $1/\beta^2$ region of Eq.~\ref{eq:bb}, where the energy loss grows dramatically. In this case, the limited dynamic range of electronics significantly limit the PID as the charge of a particle increases and the velocity decreases, leading to very large and even saturated pulses in the electronics.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. This mitigates the complete loss of information but introduces a new problem. Particles which deposit large amounts of charge will have good measurements in the low gain areas, whereas particles depositing minimal energy losses will lose information in the same low gain areas. The reconstruction of such tracks will suffer. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these issues by developing more sophisticated  pre-amplifiers and electronics \cite{feanics,feanics2}.  Never the less, it is useful to develop a software technique which extends the dynamic range of TPC electronics without the use of new external hardware. Which can be particularly useful for experiments which have already taken data with older electronics technologies. In this section we will outline a novel software technique that takes advantage of the PRF described in Section~\ref{sec:prf} and can effectively extend the dynamic range of the TPC electronics.

@@

In this case the effective dynamic range is not the single channel dynamic range, but the relative dynamic range between central pad --holding the largest charge-- and the adjacent pads --holding the smaller charges in the PRF distribution. For example, to measure minimum ionizing particles, the Signal-to-Noise-Ratio (SNR) of the pad with the smallest charge in the distribution should be some reasonable value above noise, so that the full charge distribution can be measured. From the PRF in Section~\ref{sec:prf}, we know the central pad in the cluster holds about 80\% of the total charge, where as the two adjacent pads each hold the remaining 10\%. In the \spirit TPC the electronics gain was set so that the the pads have a SNR of 6:1 for minimum ionizing tracks, and therefore the central pad had a SNR of 50:1. The maximum SNR in the central pad -- before the electronics saturate -- the SNR is 800:1. Therefore the maximum SNR is roughly 16 times larger than that of minimum ionizing particles.

@@

The Gaussian fit was performed on each PRF for a crossing angle steps of  $\ang{5}$ ranging from $\ang{0} < \theta \leq \ang{90}$. Figure~\ref{fig:normsigma} shows the two parameters resulting from fitting the the Gaussian function -- Eq.~\ref{eq:prfgaus} -- which are are plotted versus $\theta$. A $4^{th}$ order polynomial fit of these parameters allowed for interpolating for any given $\theta$ value, which is shown as the black line. Once we have this interpolation, we can predict the PRF for any given value of $\theta$.

@@

where $q_i^{\mathrm{obs}}$ are the non-saturated charges and $q_i^{\mathrm{expect}}$ are the charge values expected for that pad as calculated from the PRF, $q_i^{\mathrm{expect}} = Q\cdot PRF(\lambda_i)$. The saturated charge values $q_{i}^{\textprime}$ are treated as unknown variables and are allowed to vary in the $\chi^2$ minimization routine, they enter the calculation when they are added to get the total charge $Q = \sum q_i + \sum q_i^{'}$. The minimum $\chi^2$ value returns the best estimate for the unknown saturated charge values. The saturated hit charges are updated to reflect these estimates and the cluster position and charge is updated accordingly.

@@

To provide evidence of the success of this technique, we observe tracks which saturate pads in the high anode wire voltage region are not saturated in the low anode voltage region. While this desaturation technique avoids the need to lower the gain of any region, the low anode voltage region proved to be a direct measurement of the success of this technique.

Figure~\ref{fig:lowvshigh} shows the the $\langle dE/dx\rangle$ values of the high gain region compared with the calibrated low gain region. The effect of saturation can be seen in the high gain region for the uncorrected data above values of \SI{400}{ADC \per \milli\metre} where the values plateau, where as the low gain region still returns accurate values. Below this value the electronics are not saturated, and therefore the high and low gain sections agree. After applying the desaturation method, the correlation between the high and low gain sections is restored, as seen in Fig.~\ref{fig:lowvshigh}. From this comparison, we can approximate that the correction has corrected the high gain sections to agree with low gain sections to values of \SI{2000}{ADC \per \milli\metre}, increasing the dynamic range by a factor of a factor of at least 5.

The success of the desaturation becomes more clear when looking at the PID lines from the experimental data in Fig.~\ref{fig:data_combine}. In the following PID plots the red lines represent the most probable energy loss as given by Geant4 straggling functions. The uncorrected data in the left panel shows the effects of saturation, where the PID lines deviate significantly from their theoretical expectations starting at around \SI{400}{ADC \per \milli\metre}. After applying the desaturation technique -- in the right panel-- we see a large improvement, most notably for the He and Li particles, which suffer the most from saturation. Even the ${}^{6}Li$ and ${}^{7}Li$ particles can be separated and a more subtle improvement of the lighter particles (p, d, t), can also be seen at lower momenta. In these regions, there was little to no PID resolution before desaturation technique was applied.

@@

As the particles passes through the gas inside of the field cage, they ionize the gas creating electron-ion pairs. The drift velocities of the ions are typically \SI{e4} times slower than electron drift velocities \cite{blumrol}. Because of this, any source of ions have the potential to build up in the drift volume, creating a positive space charge. If the space charge build up is large enough, it would distort drifting electrons therefore biasing the track momentum measurement. There are several regions of the TPC in which ions are created. The largest source of positive ions is created in the avalanche process near the anode wires. But as discussed in Section~\ref{sec:wireplanes}, the gating grid captures all of the ions from this region. The other source of ions come from the primary ionization produced by the beam and reaction products in the detector gas. The energy loss  $\langle dE/dx\rangle \propto Z^2$, where Z is the charge of the particle type. Because the charge of the un-reacted beam is around Z~50, the ionization due to the beam is a factor of \num{2e3} times that of the light charged particles which mostly are of charge Z~1. Therefore the ions resulting from the un-reacted beam is the largest source of positive ions in the TPC.

@@

 The beam is positioned about \SI{25}{\centi\metre} below the anode plane and \SI{29.6}{\centi\metre} above the cathode in the TPC. It takes electrons approximately 5\si{\micro\sec} to drift to the anode plane where as it takes the ions \num{5e4}\si{\micro\sec} to drift to the cathode. The beam rate in the experiment was approximately \SI{10}{\kilo\hertz}, which has an average occurrence of 1 beam every \SI{100}{\micro\sec}. This is  much shorter than the time it takes for the ions to terminate on the cathode plane, resulting in a build up of positive ions. Figure~\ref{fig:spacechg_cartoon} gives an idea of the shape of the sheet of space charge carved out by the beam path. The ions from each beam create a line charge which drifts towards the cathode with a constant velocity. The average distance between sequential ion paths is about \SI{25}{\micro\metre} apart, therefore we expect the average number of beam paths that make up the sheet charge is around \num{1440} tracks. Since the number of tracks in the sheet charge is large, and inter beam spacing is very small, we approximate the sheet charge as a uniform charge density.

@@

This is one of the arguments for the success of the space charge correction task. Recall that only the relative distance between left and right-going tracks was minimized, and there was no guarantee that the corrected tracks coincide with the absolute BDC position at the target. The others being the agreement with the expected cocktail calibration beam as described in Section~\ref{sec:cocktail}.

>>>>

Three basic calibrations are performed on the raw data level within this task. The calibration of the electronics will be described in Section\ref{sec:elecCalib}. The anode gain calibration which will be discussed in Section~\ref{sec:anodeCalib}, and the gating grid noise subtraction and time bucket cut which is described in Section~\ref{sec:ggnoisesub}.
%gain calibration 
%ggnoise subtraction
%GG time bucket cut

\subsection{Pulse Shape Algorithm (PSA) Task}
\label{sec:psa}
%Yoffset match the TPC-Vertex_y with BDC_Y
%YPedestalOffset what is it for?
%cobo gain matching
An algorithm loops over the raw time bucket spectrum to find all the unique pulses in each pad. It is very likely that there are several pulses in a pad coming from multiple tracks passing under the same pad which are separated only in arrival time. Figure~\ref{fig:psaTask} shows the PSA task working on one particular pad in experimental data. The shaded histogram in every sub-figure represents the ADC raw spectrum of the pad. In Fig.~\ref{fig:psa0} we see the PSA has identified the first pulse and fitted it with the standard pulse; the result of the fit is shown by the red line. The fitted red line is then subtracted from the ADC spectrum, which is shown in the shaded histogram in Fig.~\ref{fig:psa1}. Here, the second pulse is found and fitted again with the same procedure; the red line will be subtracted from the raw ADC spectrum. The black line represents the sum of all the fitted pulses in each step. This continues until all pulses are found or until the PSA reaches the last bin in the spectrum. Figure~\ref{fig:psa4} shows the final result where 4 fitted pulses were found. The agreement between the total summed line and the raw ADC spectrum exemplifies how successful the PSA is in this particular set of data.

@@

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse. We define the start of the pulse to be the location in time when the rising edge of the pulse reaches of 10\% of its peak value. The height of the pulse is related to the charge induced on the pad, $Q$. Combined with the start time of the pulse $t_o$  it is saved into a new container representing a \emph{hit}, called the \emph{STHit} class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spatial coordinates, and charge information $Q$.

@@

  	\caption{Example of the pulse shape algorithm through each step in processing a given pad. }
	\label{fig:psaTask}
\end{figure}

@@

 The function of the  helix track finding algorithm is to sort the hit cloud into unique sets of hits representing tracks. This is performed by a Riemann track finding algorithm where the Cartesian space is mapped onto a Riemann sphere. This is a particular transformation in which unique helices in the Cartesian space form great circles on the Riemann sphere. This is particularly useful since hits which correspond to a track form a helix in the Cartesian space of the TPC coordinates. It is difficult to search for tracks in the hit cloud by fitting a general helix to different combinations of hits. It is much easier to search for collection of hits by fitting a plane to the great circles which are formed on the Riemann sphere. More details on the Riemann tracking can be found here \cite{spiritroot_paper}. From here the set of hits which form a unique track are stored into a new helix track container called the STHelixTrack class.

@@

\begin{figure}[!htb]
\centering
\includegraphics[scale=.5]{top_view_helix_ext.pdf}
\caption{Top down view of a fit to a track passing through several pads. Here we outline the pads in bold, where the charges $q_i$ represent the charged deposited on each pad. In this example, three clusters at the bottom are clustered in the x-direction and for the upper three clustered in the z-direction. The cluster for each group of hits represents the average position of the track. The track fit represents the position of the avalanche, $\bar{x}$,  where the position from the center to each pad to the $\bar{x}$ position is defined as $\lambda_i$.}
\label{fig:topview}
\end{figure}

 The position along the clustering direction, $\overline{X}$, is calculated as the charge-weighted average, $\overline{X} = \sum_i q_ix_i$, where $q_i$ and $x_i$ are the charge and position of the i-th hit in the cluster. The other direction is set to the center of the pad. For example, if we are clustering hits along the x-axis, the z-position is set to the center of the pad in the z-direction and vice versa. Switching the clustering direction gives better position resolution than if we clustered only along one direction. You can imagine if we calculated the clusters only along the x-axis, tracks with $\theta \approx 0^{\circ}$ the x-position is not well defined. The cluster position and charge are stored into a new container representing the cluster called the STHitCluster class. All the clusters which belong to a particular helix track is also stored in the corresponding STHelixTrack class.

@@

\subsection{Momentum and track reconstruction (GENFIT)}
After all of the corrections are applied to the clusters, the cluster's position and charge values are then passed to a software package which reconstructs the momentum of the track called GENFIT. GENFIT utilizes a Kalman fitting algorithm which returns the estimated momentum and charge values for a given track \cite{genfit}. These values are then stored into a new track container called the STRecoTrack class.

@@

\subsection{Gating grid noise subtraction}
\label{sec:ggnoisesub}
Opening the gating grid is essentially done by shorting the adjacent wires in the gating grid together, and allowing them to reach equilibrium as fast as possible. Due to coupling between the gating grid, the ground plane, and other effects, the impedance of the supply cables was not perfectly matched, causing an imbalance in the current where one side discharged faster than the other side. This caused an under-damped oscillating current in the grid, which created an oscillating residual noise early in the time bucket spectrum. Figure~\ref{fig:ggNoiseSub} shows in the upper panel shows the ADC time bucket spectrum after averaging over 2000 events in a given pad. 
 %The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency.

@@

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule
& \multicolumn{3}{c}{$100 MeV Target$}\\
\cmidrule{2-4}
Particle &\phantom{abc} & Measured & Corrected & \% Difference Raw & \% Difference Corrected\\
\midrule
p   & 882.8 & 929.5 & 877.3   &  3.7  & -1.0  \\
d   & 817.1 & 831.15 & 797.94 &  1.7  & -2.3\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail for the calibration run taken with the Aluminum target. }
\label{tb:cocktail100tar}
\end{table*}

@@

A cocktail of several light charged particles (p,d,t,${}^{3}$He,${}^{4}$He) of well defined magnetic rigidity was produced and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer were set so that the momentum resolution of the beam was defined to better than $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied with no target in place. A thick Aluminum target was also inserted to provide a slightly lower energy calibration point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases the certain particles were not able to propagate or were strongly attenuated and resulted in measurements of poor statistical precision.

@@

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The energy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. We found that the momenta of light particles in the  cocktail calibration beam differed significantly (up to 5\% ) from the expected values. This is shown in Tables~cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. We could address this discrepancy by incorporating the smaller horizontal magnetic field components at the edge of the dipole and taking into account the electron drift velocities in the direction of $\vec{E}\times\vec{B}$ as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a more general way which also includes the correction for the presence of positive space charge in the field cage volume. The same correction technique was applied here, where the cocktail beam is the special case of zero space charge.

(Tables \ref{tb:cocktail100tar} and \ref{tb:cocktail100}), protons see a slight improvement of about ~1\% whereas the deuterons are over corrected in both settings. In general, all these light particles corrected momentum are of the order of 1\% smaller than value given by the rigidity selection in the BigRIPS separator. This discrepancy could reflect the accuracy of our knowledge of the magnetic field of the SAMURAI dipole. Without a definitive explanation for this discrepancy, we have not attempted further correction of the calculated momenta. We note that it is of the same order as the resolution of the measurement of these momenta.

@@

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment, high currents being observed on the wires due to a leak in the gating grid that has since been corrected \cite{jon}. Out of all the runs used in the analysis in this thesis -- see \ref{tb:runList} -- the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we artificially increase the gain of the pads which lie above these anode wires in the software. The multiplicative factor is estimated by plotting the energy loss in the high gain sections relative to the low gain sections.

@@

Figure~\ref{fig:highlowcal} shows the correlation plot of $\langle dE/dx\rangle$ of the high versus the low gain sections. The effects of the high gain channel saturation around a value of 400 ADC\si{\per\milli\metre} and plateaus, whereas the low gain does not; this region is left out of the fit. A linear fit was performed for the calibration between the high gain channel $h_c = G\cdot l_c$, where $l_c$ is the value of the low gain channel and $G$ is the gain factor. In the case where the low anode voltage was \SI{1214}{\volt}, the gain factor was $G=9.8$.

%where in the case of \SI{1085}{\volt}, the $G = $.

 A map defining the gain calibration of each pad is pre-loaded into the software. In the decoder task the raw ADC spectra were multiplied by this gain factor for each channel. The PSA thresholds of those channels also were multiplied by the same gain factor since the noise levels were also artificially magnified as well.

@@

Using a TPC for measurements of HICs in intermediate energy heavy ion collisions presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments fundamental particles are produced with charge $\pm e$. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum ionization. In these cases the dynamic range of most TPC electronics can cover a wide range of particles. In nuclear HICs, we are interested in measured particles with charges $Ze$ where Z=1-3, and even higher in some applications. Since the energy loss in the Bethe-Bloch equation, Eq.~\ref{eq:bb}, is proportional to $Z^2$, the range of energy losses reflects the possible range of z values. Even more, in low energy nuclear HIC of intermediate beam energies -- around \SI{300}{\MeVA} -- produce low velocity particles which exist in the $1/\beta^2$ region of Eq.~\ref{eq:bb}, where the energy loss grows dramatically. In this case, the limited dynamic range of electronics significantly limit the PID as the charge of a particle increases and the velocity decreases, leading to very large and even saturated pulses in the electronics.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. This mitigates the complete loss of information but introduces a new problem. Particles which deposit large amounts of charge will have good measurements in the low gain areas, whereas particles depositing minimal energy losses will lose information in the same low gain areas. The reconstruction of such tracks will suffer. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these issues by developing more sophisticated  pre-amplifiers and electronics \cite{feanics,feanics2}.  Nevertheless, it is useful to develop a software technique which extends the dynamic range of TPC electronics without the use of new external hardware. Which can be particularly useful for experiments which have already taken data with older electronics technologies. In this section we will outline a novel software technique that takes advantage of the PRF described in Section~\ref{sec:prf} and can effectively extend the dynamic range of the TPC electronics.

@@

In this case the effective dynamic range is not the single channel dynamic range, but the relative dynamic range between central pad --holding the largest charge-- and the adjacent pads --holding the smaller charges in the PRF distribution. For example, to measure minimum ionizing particles, the Signal-to-Noise-Ratio (SNR) of the pad with the smallest charge in the distribution should be some reasonable value, so that the full charge distribution can be measured. From the PRF in Section~\ref{sec:prf}, we know the central pad in the cluster holds about 80\% of the total charge, whereas the two adjacent pads each hold the remaining 10\%. In the \spirit TPC the electronics gain was set so that the the pads have a SNR of 6:1 for minimum ionizing tracks, and therefore the central pad had a SNR of 50:1. The maximum SNR in the central pad -- before the electronics saturate -- the SNR is 800:1. Therefore the maximum SNR is roughly 16 times larger than that of minimum ionizing particles.

@@

The Gaussian fit was performed on each PRF for a crossing angle steps of  $\ang{5}$ ranging from $\ang{0} < \theta \leq \ang{90}$. Figure~\ref{fig:normsigma} shows the two parameters resulting from fitting the the Gaussian function -- Eq.~\ref{eq:prfgaus} -- which are plotted versus $\theta$. A $4^{th}$ order polynomial fit of these parameters allowed for interpolating for any given $\theta$ value, which is shown as the black line. Once we have this interpolation, we can predict the PRF for any given value of $\theta$.

@@

where $q_i^{\mathrm{obs}}$ are the non-saturated charges and $q_i^{\mathrm{expect}}$ are the charge values expected for that pad as calculated from the PRF, $q_i^{\mathrm{expect}} = Q\cdot PRF(\lambda_i)$. The saturated charge values $q_{i}^{\textprime}$ are treated as unknown variables and are allowed to vary in the $\chi^2$ minimization routine, they enter the calculation when they are added to get the total charge $Q = \sum q_i + \sum q_i^{'}$. The minimum $\chi^2$ value returns the best estimate for the unknown saturated charge values. The saturated hit charges are updated to reflect these estimates and the cluster position and charge are updated accordingly.

@@

To provide evidence of the success of this technique, we observe tracks which saturate pads in the high anode wire voltage region but are not saturated in the low anode voltage region. While this desaturation technique avoids the need to lower the gain of any region, the low anode voltage region proved to be a direct measurement of the success of this technique.

Figure~\ref{fig:lowvshigh} shows the the $\langle dE/dx\rangle$ values of the high gain region compared with the calibrated low gain region. The effect of saturation can be seen in the high gain region for the uncorrected data above values of \SI{400}{ADC \per \milli\metre} where the values plateau, whereas the low gain region still returns accurate values. Below this value the electronics are not saturated, and therefore the high and low gain sections agree. After applying the desaturation method, the correlation between the high and low gain sections is restored, as seen in Fig.~\ref{fig:lowvshigh}. From this comparison, we can approximate that the correction has corrected the high gain sections to agree with low gain sections to values of \SI{2000}{ADC \per \milli\metre}, increasing the dynamic range by a factor of a factor of at least 5.

The success of the desaturation becomes more clear when looking at the PID lines from the experimental data in Fig.~\ref{fig:data_combine}. In the following PID plots the red lines represent the most probable energy loss as given by Geant4 straggling functions. The uncorrected data in the left panel shows the effects of saturation, where the PID lines deviate significantly from their theoretical expectations starting at around \SI{400}{ADC \per \milli\metre}. After applying the desaturation technique -- in the right panel-- we see a large improvement, most notably for the He and Li particles, which suffer the most from saturation. Even the ${}^{6}$Li and ${}^{7}$Li particles can be separated and a more subtle improvement of the lighter particles (p, d, t), can also be seen at lower momenta. In these regions, there was little to no PID resolution before desaturation technique was applied.

@@

As the particles pas through the gas inside of the field cage, they ionize the gas creating electron-ion pairs. The drift velocities of the ions are typically \SI{e4} times slower than electron drift velocities \cite{blumrol}. Because of this, any source of ions have the potential to build up in the drift volume, creating a positive space charge. If the space charge build up is large enough, it would distort drifting electrons therefore biasing the track momentum measurement. There are several regions of the TPC in which ions are created. The largest source of positive ions is created in the avalanche process near the anode wires. But as discussed in Section~\ref{sec:wireplanes}, the gating grid captures all of the ions from this region. The other source of ions comes from the primary ionization produced by the beam and reaction products in the detector gas. The energy loss  $\langle dE/dx\rangle \propto Z^2$, where Z is the charge of the particle type. Because the charge of the un-reacted beam is around Z=50, the ionization due to the beam is a factor of \num{2e3} times that of the light charged particles which mostly are of charge Z=1. Therefore the ions resulting from the un-reacted beam is the largest source of positive ions in the TPC.

@@

 The beam is positioned about \SI{25}{\centi\metre} below the anode plane and \SI{29.6}{\centi\metre} above the cathode in the TPC. It takes electrons approximately 5\si{\micro\sec} to drift to the anode plane whereas it takes the ions \num{5e4}\si{\micro\sec} to drift to the cathode. The beam rate in the experiment was approximately \SI{10}{\kilo\hertz}, which has an average occurrence of 1 beam every \SI{100}{\micro\sec}. This is  much shorter than the time it takes for the ions to terminate on the cathode plane, resulting in a build up of positive ions. Figure~\ref{fig:spacechg_cartoon} gives an idea of the shape of the sheet of space charge carved out by the beam path. The ions from each beam create a line charge which drifts towards the cathode with a constant velocity. The average distance between sequential ion paths is about \SI{25}{\micro\metre} apart, therefore we expect the average number of beam paths that make up the sheet charge is around \num{1440} tracks. Since the number of tracks in the sheet charge is large, and inter beam spacing is very small, we approximate the sheet charge as a uniform charge density.

@@

This is one of the arguments for the success of the space charge correction task. Recall that only the relative distance between left and right-going tracks was minimized, and there was no guarantee that the corrected tracks coincide with the absolute BDC position at the target. The others being in agreement with the expected cocktail calibration beam as described in Section~\ref{sec:cocktail}.
